---
title: "Untitled"
output: html_document
date: "2024-08-03"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(RColorBrewer)
library(patchwork)
library(ComplexHeatmap)
library(grid)
library(circlize)

library(hdf5r)
library(arrow)
library(Seurat)
library(presto)

library(clusterProfiler)
library(org.Mm.eg.db)

dir.s <- "Setup"
dir.r <- "Results"
dir.mic <- "Results/Microenvironment"
dir.mg <- "Results/Microenvironment/GSEA"

cluster_DE_wt <- readRDS(paste0(dir.mic, "/cluster_DE_wt.RDS"))
WT <- readRDS(file.path(dir.s, "WT.spatial_annotated.rds"))

colors.mac <- c(
  "1" = "#E76F51",
  "2" = "#E9C46A",
  "3" = "#36BA98",
  "4" = "#03396c"
)
```


```{r setup, include=FALSE}
new_metadata <- as.character(WT@meta.data$MAC_type_filtered)

table(WT@meta.data$MAC_type_filtered, useNA="ifany")
# 3. Assign the names of the new metadata vector to be the Seurat cell names.
names(new_metadata) <- rownames(WT@meta.data)

# Optionally check the head to verify the names are now cell barcodes:
head(new_metadata)

# 4. Add the new metadata column to the Seurat object with AddMetaData()
WT <- AddMetaData(WT, metadata = new_metadata, col.name = "MAC_type_broad")


colnames(WT@meta.data)
table(WT$MAC_type_broad, useNA="ifany")
table(WT$first_type, useNA="ifany")
```

#named colors for cell_type
```{r}
levels <- c("FB", "CHFB", "PVC", "KRT", "LEC", "APC-MAC", "LYM", "NEU", "SCH", "VEC", "NA")

all_colors <- c("orange", "#1F78B4", "pink", "#66A61E", "#005F6A",
                "brown", "#666666", "#A6761D", "yellow", "#6A3D9A", "#BDBDBD")

colors.celltype <- setNames(all_colors, levels)
```

#add hires image
```{r}
WT@images[["slice1"]] <- Read10X_Image(
  image.dir = "Q:/IFI-F-AG-Henneke-Forschung/Mitarbeiter/Florens/visium_HD/re-run/Spaceranger/WT2-auto/outs/binned_outputs/square_008um/Spatial/", image.name = "tissue_hires_image.png")
```

#Working: "Find neighbours to each spot of a specific cluster WT
#First, correlate radius to number of neighbor spots
```{r}
radius <- 50

calculate_average_neighbors_dual <- function(coord_df, radius, meta_data,
                                             center_field = "MAC_type_broad",
                                             neighbor_field = "first_type") {
  # Centers: MAC_type_broad != NA
  center_spots <- rownames(meta_data)[!is.na(meta_data[[center_field]])]
  center_coords <- coord_df[coord_df$spot %in% center_spots, , drop = FALSE]
  
  # Neighbor pools
  neighbor_spots_ft <- rownames(meta_data)[!is.na(meta_data[[neighbor_field]])]
  neighbor_coords_ft <- coord_df[coord_df$spot %in% neighbor_spots_ft, , drop = FALSE]
  neighbor_coords_all <- coord_df
  
  if (nrow(center_coords) == 0) return(c(avg_first_type = NA_real_, avg_all = NA_real_))
  if (nrow(neighbor_coords_ft) == 0) avg_first_type <- NA_real_ else {
    avg_first_type <- mean(sapply(center_coords$spot, function(spot) {
      target <- center_coords[center_coords$spot == spot, c("x","y"), drop = FALSE]
      d <- sqrt((neighbor_coords_ft$x - target$x)^2 + (neighbor_coords_ft$y - target$y)^2)
      sum(d <= radius & neighbor_coords_ft$spot != spot)
    }))
  }
  
  avg_all <- mean(sapply(center_coords$spot, function(spot) {
    target <- center_coords[center_coords$spot == spot, c("x","y"), drop = FALSE]
    d <- sqrt((neighbor_coords_all$x - target$x)^2 + (neighbor_coords_all$y - target$y)^2)
    sum(d <= radius & neighbor_coords_all$spot != spot)
  }))
  
  c(avg_first_type = avg_first_type, avg_all = avg_all)
}

# Prepare inputs
coordinates <- GetTissueCoordinates(WT)
coord_df <- as.data.frame(coordinates)
coord_df$spot <- rownames(coord_df)
meta_data <- WT@meta.data

# Run
avg_neighbors_both <- calculate_average_neighbors_dual(coord_df, radius, meta_data)

cat(
  "Average neighbors within", radius, "\n",
  "  - using centers MAC_type_broad != NA and neighbors first_type != NA:", avg_neighbors_both["avg_first_type"], "\n",
  "  - using centers MAC_type_broad != NA and neighbors ANY spot:", avg_neighbors_both["avg_all"], "\n"
)
```



```{r}

radius <- 50  

find_neighbors <- function(spot, target_coords, coord_df, radius) {
  target <- target_coords[spot, c("x", "y")]
  distances <- sqrt((coord_df$x - target$x)^2 + (coord_df$y - target$y)^2)
  nearby_spots <- coord_df$spot[distances <= radius & coord_df$spot != spot]  
  return(nearby_spots)
}

# Get coordinates 
coordinates <- GetTissueCoordinates(WT)
coord_df <- as.data.frame(coordinates)
coord_df$spot <- rownames(coord_df)

cluster_DE_wt <- list()

for (cluster_id in 1:4) {
  
  # Step 1: Identify spots in the current cluster
  cluster_spots <- WhichCells(WT, expression = MAC_type_broad == cluster_id)
  
  target_coords <- coord_df %>% filter(spot %in% cluster_spots)
  
  # Step 2: Find neighbors for each spot in the current cluster
  neighbors <- lapply(cluster_spots, find_neighbors, target_coords = target_coords, coord_df = coord_df, radius = radius)
  names(neighbors) <- cluster_spots
  
  # Combine all neighborhood spots, excluding each original spot
  all_neighbors <- unique(unlist(lapply(names(neighbors), function(spot) {
    setdiff(neighbors[[spot]], spot)
  })))
  
  # Step 3: Set identities for DE analysis
  Idents(WT) <- "original"  
  WT <- SetIdent(WT, cells = all_neighbors, value = paste0("neighbors_cluster_", cluster_id))
  WT <- SetIdent(WT, cells = setdiff(Cells(WT), all_neighbors), value = "other")
  
  # Step 4: Run FindMarkers for the current cluster's neighborhood
  DE_results <- FindMarkers(WT, ident.1 = paste0("neighbors_cluster_", cluster_id),  ident.2 = "other", only.pos=T, assay = "Spatial")
  
  # Order DE results by avg_log2FC and store in the list
  DE_results <- DE_results[order(DE_results$avg_log2FC, decreasing = TRUE), ]
  cluster_DE_wt[[paste0("cluster_", cluster_id)]] <- DE_results
}

head(cluster_DE_wt[["cluster_1"]], n=10)
length(rownames(cluster_DE_wt[["cluster_1"]]))

saveRDS(cluster_DE_wt, paste0(dir.mic, "/cluster_DE_wt.RDS"))

```

```{r}
neighbors_by_cluster <- list()

for (cluster_id in 1:4) {
  # Step 1: Identify spots in the current cluster
  cluster_spots <- WhichCells(WT, expression = MAC_type_broad == cluster_id)
  
  target_coords <- coord_df %>% filter(spot %in% cluster_spots)
  
  # Step 2: Find neighbors for each spot in the current cluster
  neighbors <- lapply(cluster_spots, find_neighbors, target_coords = target_coords, coord_df = coord_df, radius = radius)
  names(neighbors) <- cluster_spots
  
  # Combine all neighborhood spots, excluding each original spot
  all_neighbors <- unique(unlist(lapply(names(neighbors), function(spot) {
    setdiff(neighbors[[spot]], spot)
  })))
  
  neighbors_by_cluster[[paste0("cluster_", cluster_id)]] <- all_neighbors
}

head(neighbors_by_cluster[["cluster_1"]], n=10)
length(neighbors_by_cluster[["cluster_4"]])

saveRDS(neighbors_by_cluster, file = paste0(dir.mic, "/neighbors_by_cluster.rds"))
```

#Heatmap neighbors
```{r}
neighbors_by_cluster <- readRDS(file = paste0(dir.mic, "/neighbors_by_cluster.rds"))
```

#Heatmap neighbors data preparation
```{r}
top_n <- 8  

# Initialize a list
top_deg_list <- list()

# Extract top DEGs for each cluster
for (cluster_id in 1:4) {
  cluster_key <- paste0("cluster_", cluster_id)
  
  if (!is.null(cluster_DE_wt[[cluster_key]]) && nrow(cluster_DE_wt[[cluster_key]]) > 0) {
    # Extract the top DEGs based on avg_log2FC
    top_deg_list[[cluster_key]] <- rownames(cluster_DE_wt[[cluster_key]][1:min(top_n, nrow(cluster_DE_wt[[cluster_key]])), ])
  }
}

# Combine top DEGs from all clusters into a unique set
all_top_genes <- unique(unlist(top_deg_list))

# Retrieve normalized expression data from WT
expression_data <- GetAssayData(WT, layer  = "data", assay = "Spatial")

# Subset expression data for the top DEGs
heatmap_expression <- expression_data[all_top_genes, , drop = FALSE]


# Calculate average expression for each cluster neighborhood
avg_expression <- sapply(1:4, function(cluster_id) {
  cluster_neighbors <- neighbors_by_cluster[[paste0("cluster_", cluster_id)]]
  
  # Ensure valid neighbors
  valid_neighbors <- cluster_neighbors[cluster_neighbors %in% colnames(heatmap_expression)]
  
  if (length(valid_neighbors) == 0) {
    return(rep(NA, nrow(heatmap_expression)))  
  }
  
  # Calculate rowMeans for valid neighbors
  rowMeans(heatmap_expression[, valid_neighbors, drop = FALSE])
})

# Assign column names for clarity
colnames(avg_expression) <- paste0("Cluster ", 1:4)

head(avg_expression)

# Scale the data (row-wise scaling)
avg_expression_scaled <- t(scale(t(avg_expression)))
```

#Heatmap neighbors visualization
```{r}
# Prepare gene labels in italic
italic_gene_labels <- paste0("italic('", rownames(avg_expression_scaled), "')")

cols <- colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(101)

# Plot the heatmap
png(paste0(dir.mic, "/heatmap_top_DEG_neighbors.png"), width = 1700, height = 3000, res = 600)

Heatmap(
  avg_expression_scaled,
  name = "Z-score",            
  col = cols, 
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  row_labels = parse(text = italic_gene_labels),  
  column_names_rot = 45
)

dev.off()

```


#Perform GSEA in neighbor DEGs and store as separate files
```{r}
perform_gsea <- function(cluster_DE_results, orgdb, num_clusters = 4, dataset_name = "default") {
  
  perform_gsea_for_cluster <- function(cluster_data, cluster_id) {
    set.seed(123) 
    
    cluster_data$gene <- rownames(cluster_data)
    rownames(cluster_data) <- NULL
    
    eg <- bitr(cluster_data$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = orgdb)
    eg <- subset(eg, !duplicated(eg$SYMBOL))
    rownames(eg) <- eg$SYMBOL
    eg$SYMBOL <- NULL
  
    diffexpgenes <- merge(cluster_data, eg, by.x = "gene", by.y = "row.names")
    rownames(diffexpgenes) <- diffexpgenes$gene
    diffexpgenes$gene <- NULL
 
    fc <- diffexpgenes$avg_log2FC
    names(fc) <- diffexpgenes$ENTREZID
    fc <- sort(fc, decreasing = TRUE)
    
    gse <- gseGO(geneList = fc, ont = "BP", minGSSize = 5, maxGSSize = 800, OrgDb = orgdb, seed = 1)
    
    if (is.null(gse) || nrow(gse@result) == 0) {
      message(paste("No significant GO terms for cluster", cluster_id, "in dataset", dataset_name))
      return(NULL)
    }
    
    csv_output <- file.path(dir.mg, paste0(dataset_name, "_gse_cluster_", cluster_id, ".csv"))
    rds_output <- file.path(dir.mg, paste0(dataset_name, "_gse_cluster_", cluster_id, ".rds"))
    
    write.csv(gse@result, csv_output, row.names = FALSE)
    saveRDS(gse, rds_output)
    
    message(paste("GSEA results saved for cluster", cluster_id, "in dataset", dataset_name, "at", csv_output, "and", rds_output))
  }
  
  for (cluster_id in 1:num_clusters) {

    cluster_key <- paste0("cluster_", cluster_id)
    if (!cluster_key %in% names(cluster_DE_results)) {
      message(paste("Cluster", cluster_id, "not found in cluster_DE_results for dataset", dataset_name))
      next
    }
    
    cluster_data <- cluster_DE_results[[cluster_key]]
    
    if (is.null(cluster_data) || nrow(cluster_data) == 0) {
      message(paste("No DE genes available for cluster", cluster_id, "in dataset", dataset_name))
      next
    }
    
    perform_gsea_for_cluster(cluster_data, cluster_id)
  }
}

perform_gsea(cluster_DE_wt, orgdb = org.Mm.eg.db, num_clusters = 4, dataset_name = "wt")
```

#use simplify on gseas
```{r}
simplified_gse_results_wt <- list()

for (cluster_id in 1:4) {
  gse_file <- file.path(dir.mg, paste0("wt_gse_cluster_", cluster_id, ".rds"))
  
  gse_cluster <- readRDS(gse_file)
  
  simplified <- clusterProfiler::simplify(
    gse_cluster,
    cutoff = 0.3,  
    by = "p.adjust",
    select_fun = min,
    measure = "Wang"
  )
  
  # Store in list
  simplified_gse_results_wt[[paste0("cluster_", cluster_id)]] <- simplified
}
```

#select unique top terms
```{r}
all_top_terms_wt <- c()

for (cluster_id in 1:4) {
  
  cluster_results <- simplified_gse_results_wt[[paste0("cluster_", cluster_id)]]
  
  if (is.null(cluster_results)) {
    message(paste("No results for cluster", cluster_id))
    next
  }
  
  significant_terms <- cluster_results@result %>%
    filter(p.adjust < 0.05) %>%  
    arrange(desc(abs(NES)))      
  
  # Extract the top terms based on NES
  top_terms <- head(significant_terms$Description, 50)

  # Append 
  all_top_terms_wt <- c(all_top_terms_wt, top_terms)
}

# Create a list of unique terms across all clusters
unique_top_terms_wt <- unique(all_top_terms_wt)

message("Unique top terms across all clusters:")
print(unique_top_terms_wt)
```


## Capitalize first letter
```{r}
capitalize_first <- function(string) {
  paste0(toupper(substr(string, 1, 1)), substr(string, 2, nchar(string)))
}
```

## process and plot the data
#Comment: Big overlap between clusters
```{r}
generate_enrichment_plot <- function(unique_top_terms, simplified_gse_results, clusters, output_file, dataset_name) {
  library(dplyr)
  
  unique_top_terms <- sapply(unique_top_terms, capitalize_first)
  unique_top_terms <- unique(unique_top_terms)  

  enrich_df_select_all <- data.frame()
  
  for (cluster_id in clusters) {
    cluster_results <- simplified_gse_results[[paste0("cluster_", cluster_id)]]
    
    if (is.null(cluster_results)) {
      next
    }
    
    # Capitalize Description
    cluster_results@result$Description <- sapply(cluster_results@result$Description, capitalize_first)
    
    # Filter for terms of interest
    selected_results <- cluster_results@result %>%
      filter(Description %in% unique_top_terms) %>%
      mutate(
        Condition = paste0("Cluster_", cluster_id),
        Description_short = substr(Description, 1, 50)
      )
    
    enrich_df_select_all <- bind_rows(enrich_df_select_all, selected_results)
  }
  
  for (missing_term in setdiff(unique_top_terms, enrich_df_select_all$Description)) {
    if (!is.na(missing_term)) {  
      for (cluster_id in clusters) {
        enrich_df_select_all <- bind_rows(enrich_df_select_all, data.frame(
          Description = missing_term,
          p.adjust = NA,
          NES = NA,  
          Condition = paste0("Cluster_", cluster_id),
          Description_short = substr(missing_term, 1, 50),
          stringsAsFactors = FALSE
        ))
      }
    }
  }
  
  # Remove rows with NA in Description
  enrich_df_select_all <- enrich_df_select_all %>%
    filter(!is.na(Description) & Description != "NA")
  
  # Explicitly remove NA from factor levels in Description_short
  enrich_df_select_all$Description_short <- factor(enrich_df_select_all$Description_short)
  enrich_df_select_all$Description_short <- droplevels(enrich_df_select_all$Description_short)
  
  # Replace underscores in cluster labels
  enrich_df_select_all$Condition <- gsub("_", " ", enrich_df_select_all$Condition)
  
  # Ensure `Condition` is a factor
  enrich_df_select_all$Condition <- factor(enrich_df_select_all$Condition, levels = paste("Cluster", clusters))
  
  # Plot the heatmap 
  plot <- ggplot(enrich_df_select_all, aes(Condition, Description_short, fill = NES)) + 
    geom_tile() +
    scale_fill_gradientn(
      name = "Enrichment", 
      colors = rev(brewer.pal(n = 11, name = "RdYlBu")),
      na.value = "grey90"
    ) +
    theme_classic() +
    labs(y = "", x = "") +
    theme(
      panel.background = element_rect(fill = "grey", colour = "grey", linewidth = 0.5, linetype = "solid"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      axis.text.y = element_text(size = 12),
      legend.position = "bottom"
    )
  
  print(plot)
  
  ggsave(output_file, plot = plot, width = 6, height = 6, dpi = 600)
}
```


# Generate the plot
```{r}
generate_enrichment_plot(
  unique_top_terms = unique_top_terms_wt,
  simplified_gse_results = simplified_gse_results_wt,
  clusters = 1:4,
  output_file = paste0(dir.mic, "/enrichGO_wt.png"),
  dataset_name = "WT"
)
```



# So instead identify distinguishing terms
```{r}
find_distinguishing_terms <- function(simplified_gse_results, clusters, p_cutoff = 0.05, top_n) {
  library(dplyr)
  
  nes_data <- data.frame()
  
  for (cluster_id in clusters) {
    cluster_results <- simplified_gse_results[[paste0("cluster_", cluster_id)]]
    
    if (is.null(cluster_results)) {
      next
    }
    
    # Filter for significant terms
    significant_terms <- cluster_results@result %>%
      filter(p.adjust < p_cutoff) %>%
      dplyr::select(Description, NES) %>%
      mutate(Cluster = paste0("Cluster_", cluster_id))
    
    nes_data <- bind_rows(nes_data, significant_terms)
  }
  
  # Pivot NES data into a wide format
  nes_wide <- nes_data %>%
    pivot_wider(names_from = Cluster, values_from = NES, values_fill = NA)
  
  # Calculate standard deviation of NES across clusters
  nes_wide <- nes_wide %>%
    rowwise() %>%
    mutate(NES_sd = sd(c_across(starts_with("Cluster_")), na.rm = TRUE)) %>%
    ungroup()
  
  # Rank terms by variability and select the top N
  top_distinguishing_terms <- nes_wide %>%
    arrange(desc(NES_sd)) %>%
    slice_head(n = top_n)
  
  return(top_distinguishing_terms)
}

distinguishing_terms <- find_distinguishing_terms(
  simplified_gse_results = simplified_gse_results_wt,
  clusters = 0:8,
  p_cutoff = 0.05,
  top_n = 50
)

print(distinguishing_terms)

```


#plot distinguishing terms
#comment: Terms still very similar despite simplify but better to differentiate between clusters.
```{r}
heatmap_data <- distinguishing_terms %>%
  dplyr::select(-NES_sd) %>%  
  pivot_longer(
    cols = starts_with("Cluster_"),  
    names_to = "Condition",
    values_to = "NES"
  ) %>%
  mutate(
    Description_short = substr(Description, 1, 50),  
    Condition = gsub("_", " ", Condition)  
  )

heatmap_data$Condition <- factor(heatmap_data$Condition, levels = unique(heatmap_data$Condition))

# Plot the heatmap
plot <- ggplot(heatmap_data, aes(Condition, Description_short, fill = NES)) + 
  geom_tile() +
  scale_fill_gradientn(
    name = "Enrichment", 
    colors = rev(brewer.pal(n = 11, name = "RdYlBu")),
    na.value = "grey90"
  ) +
  theme_classic() +
  labs(y = "Distinguishing GO Terms", x = "Clusters") +
  theme(
    panel.background = element_rect(fill = "grey", colour = "grey", linewidth = 0.5, linetype = "solid"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "bottom"
  )

# Print the plot
print(plot)


```

##short list unique terms
```{r}

View(simplified_gse_results_wt$cluster_3@result)

go_terms <- c(	
"regulation of lymphocyte proliferation", 
"regulation of T cell proliferation",
"positive regulation of cytokine production",
"proton motive force-driven ATP synthesis",
"leukocyte chemotaxis"
)

```

#plot shortlisted terms
```{r}
generate_enrichment_plot_manual <- function(go_terms, simplified_gse_results, clusters, output_file, dataset_name) {

  go_terms <- sapply(go_terms, capitalize_first)
  go_terms <- unique(go_terms)  

  enrich_df_select_all <- data.frame()
  
  for (cluster_id in clusters) {
    cluster_results <- simplified_gse_results[[paste0("cluster_", cluster_id)]]
    
    if (is.null(cluster_results)) {
      next
    }
    
    # Capitalize Description 
    cluster_results@result$Description <- sapply(cluster_results@result$Description, capitalize_first)
    
    # Filter for manually provided GO
    selected_results <- cluster_results@result %>%
      filter(Description %in% go_terms) %>%
      mutate(
        Condition = paste0("Cluster_", cluster_id),
        Description_short = substr(Description, 1, 50)  
      )
    
    enrich_df_select_all <- bind_rows(enrich_df_select_all, selected_results)
  }
  
  
  for (missing_term in setdiff(go_terms, enrich_df_select_all$Description)) {
    if (!is.na(missing_term)) {  
      for (cluster_id in clusters) {
        enrich_df_select_all <- bind_rows(enrich_df_select_all, data.frame(
          Description = missing_term,
          p.adjust = NA,
          NES = NA,  
          Condition = paste0("Cluster_", cluster_id),
          Description_short = substr(missing_term, 1, 50),
          stringsAsFactors = FALSE
        ))
      }
    }
  }
  
  enrich_df_select_all <- enrich_df_select_all %>%
    filter(!is.na(Description) & Description != "NA")
  
  enrich_df_select_all$Description_short <- factor(enrich_df_select_all$Description_short)
  enrich_df_select_all$Description_short <- droplevels(enrich_df_select_all$Description_short)
  
  # Replace underscores
  enrich_df_select_all$Condition <- gsub("_", " ", enrich_df_select_all$Condition)
  
  enrich_df_select_all$Condition <- factor(enrich_df_select_all$Condition, levels = paste("Cluster", clusters))
  
  # Plot the heatmap 
  plot <- ggplot(enrich_df_select_all, aes(Condition, Description_short, fill = NES)) + 
    geom_tile() +
    scale_fill_gradientn(
      name = "NES", 
      colors = rev(brewer.pal(n = 11, name = "RdYlBu")),
      na.value = "grey90"
    ) +
    theme_classic() +
    labs(y = "", x = "") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      axis.text.y = element_text(size = 12),
      legend.position = "bottom"
    )
  
  print(plot)
  
  ggsave(output_file, plot = plot, width = 5, height = 2.4, dpi = 600)
}


#panel.background = element_rect(fill = "white", colour = "grey", linewidth = 0.5, linetype = "solid"),
```

```{r}
generate_enrichment_plot_manual(
  go_terms = go_terms,
  simplified_gse_results = simplified_gse_results_wt,
  clusters = 1:4,
  output_file = paste0(dir.mic, "/manual_enrichment_heatmap.png"),
  dataset_name = "WT"
)
```




#BuildNicheAssay quantifies neighboring cells as identified by clusters.
```{r}
table(WT@meta.data$MAC_type_broad, useNA="ifany")

Idents(WT) <- "first_type"

WT_filtered <- subset(WT, cells = colnames(WT)[!is.na(WT$first_type)])

WT_niche <- BuildNicheAssay(
  object = WT_filtered,
  fov= "slice1",
  assay = "Niche",
  neighbors.k = 30,
  niches.k = 4,
  group.by = "first_type"
)

# Fetch enrichment data 
niche_data <- GetAssayData(WT_niche, assay = "Niche", layer = "data")

dim(niche_data)  
head(niche_data)

# Fetch metadata 
meta_data <- WT_niche@meta.data
spot_types <- meta_data$first_type  


valid_spots <- !is.na(spot_types)
niche_data <- niche_data[, valid_spots]  
spot_types <- spot_types[valid_spots]   

dim(niche_data)
table(spot_types)

# Calculate mean enrichment scores 
niche_by_type <- sapply(unique(spot_types), function(type) {
  spots_of_type <- which(spot_types == type)  
  rowMeans(niche_data[, spots_of_type, drop = FALSE])  
})

niche_by_type <- as.matrix(niche_by_type)

# Assign row and column names
rownames(niche_by_type) <- rownames(niche_data)  
colnames(niche_by_type) <- unique(spot_types)    

head(niche_by_type)

niche_by_type_scaled <- t(scale(t(niche_by_type)))

# Check scaled matrix
head(niche_by_type_scaled)


# Define the desired order 
cell_type_order <- rownames(niche_by_type_scaled)  

# Reorder 
niche_by_type_ordered <- niche_by_type_scaled[cell_type_order, cell_type_order]

output_file <- "Results/Microenvironment/niche_enrichment_heatmap.png"

cols <- colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(101)

png(filename = output_file, width = 2000, height = 1800, res = 600)

# Important! Row is the central cell type, col is enriched type! So mac enriched around lym

Heatmap(
  niche_by_type_ordered,
  name="Z-score",
  cluster_rows  = FALSE,
  cluster_columns  = FALSE,
  col  = cols
)

dev.off()

```

```{r}
#create enrichment of cell types heatmap

set.seed(123)  # ensure determinism

safe_scale_row <- function(v) {
  m <- mean(v, na.rm = TRUE); s <- sd(v, na.rm = TRUE)
  if (!is.finite(s) || s == 0) return(rep(0, length(v)))
  (v - m) / s
}

k <- 15

# 1) Build Niche assay
Mac_niche <- BuildNicheAssay(
  object      = WT_filtered,
  fov         = "slice1",
  assay       = "Niche",
  neighbors.k = k,
  niches.k    = 4,
  group.by    = "first_type"
)

# 2) Matrix & metadata
niche_data <- GetAssayData(Mac_niche, assay = "Niche", layer = "data")
spot_types <- Mac_niche@meta.data$MAC_type_broad

# 3) Keep valid MAC spots
valid_spots <- !is.na(spot_types)
niche_data  <- niche_data[, valid_spots, drop = FALSE]
spot_types  <- spot_types[valid_spots]

# 4) Aggregate by central MAC type
mac_levels <- unique(spot_types)
niche_by_first_around_mac <- sapply(mac_levels, function(mac) {
  cols_idx <- which(spot_types == mac)
  rowMeans(niche_data[, cols_idx, drop = FALSE])
})
niche_by_first_around_mac <- as.matrix(niche_by_first_around_mac)
rownames(niche_by_first_around_mac) <- rownames(niche_data)
colnames(niche_by_first_around_mac) <- mac_levels

# 5) Drop zero-variance rows (match loop), then safe z-score
row_sd <- apply(niche_by_first_around_mac, 1, sd, na.rm = TRUE)
keep_rows <- is.finite(row_sd) & row_sd > 0
niche_var <- niche_by_first_around_mac[keep_rows, , drop = FALSE]

niche_scaled <- t(apply(niche_var, 1, safe_scale_row))
niche_scaled[!is.finite(niche_scaled)] <- 0

desired_cols <- as.character(1:4)
col_order <- intersect(desired_cols, colnames(niche_scaled))
niche_scaled <- niche_scaled[, col_order, drop = FALSE]

# Map from original names -> display labels
lab_map <- setNames(paste("Cluster", desired_cols), desired_cols)
display_labels <- lab_map[colnames(niche_scaled)]


# 8) Plot (same options as loop)
output_file <- "Results/Microenvironment/niche_enrichment_celltypes_around_MACtypes_k15.png"
dir.create(dirname(output_file), showWarnings = FALSE, recursive = TRUE)
png(filename = output_file, width = 1700, height = 1700, res = 600)

Heatmap(
  niche_scaled,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  col = cols,
  column_names_rot = 45,
  na_col = "grey90",
  heatmap_legend_param = list(at = c(-2, -1, 0, 1, 2)),
  column_labels = display_labels
)

dev.off()

```



