---
title: "Untitled"
output: html_document
date: "2024-08-03"
editor_options: 
  chunk_output_type: console
---




```{r setup, include=FALSE}
library(hdf5r)
library(arrow)
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(spacexr)
library(parallel)
library(Matrix)
library(tibble)
library(biomaRt)

dir.s <- "Setup"
dir.r <- "Results"

colors.mac <- c("#8AB6F9",  "#ffc13b","#04d4f0", "#5c3c92", "#077b8a", "#d72631",  "#80c904")

```

# read objects
```{r}
WT_annotated <- readRDS(paste0(dir.s, "/WT.spatial_annotated.rds"))

integrated <- readRDS(paste0(dir.s, "/integrated.RDS"))

mac_sc <- readRDS(paste0(dir.s, "/mac_filter_merge.RDS"))
mac_sc@meta.data$Experiment2
mac_sc_wt <- subset(mac_sc, subset = Experiment2 == "WT")
table(mac_sc_wt@meta.data$new_clusters)

mac_sc_wt$broad_clusters <- ifelse(grepl("^4", as.character(mac_sc_wt$new_clusters)), 
                                   "4", 
                                   as.character(mac_sc_wt$new_clusters))

# Make it an ordered factor (optional)
mac_sc_wt$broad_clusters <- factor(mac_sc_wt$broad_clusters, levels = c("1","2","3","4"))

# Quick check
table(mac_sc_wt$broad_clusters, useNA = "ifany")
```

#read files 8µm
```{r}
WT <- Load10X_Spatial(paste0(dir.s, "/square_008um"))

WT <- NormalizeData(WT)
```

#tech visualization
```{r}
vln.plot.w <- VlnPlot(WT, features = "nCount_Spatial", pt.size = 0) + theme(axis.text = element_text(size = 4)) + NoLegend()

count.plot.w <- SpatialFeaturePlot(WT, features = "nCount_Spatial") + theme(legend.position = "right")

vln.plot.w | count.plot.w
ggsave(paste0(dir.s, "/counts.wt.png"), width=6, height=3, dpi=600)

VlnPlot(WT, features = "nFeature_Spatial", log=T)
total_features <-nrow(WT@assays$Spatial@layers$counts)

print(total_features)
total_spots <- ncol(WT@assays$Spatial@layers$counts)
print(total_spots)

```

#cluster the whole dataset WT
```{r}
set.seed(123)

min_count <- 50  

WT <- subset(WT, subset = nCount_Spatial > min_count)

WT <- FindVariableFeatures(WT)

WT <- ScaleData(WT)

WT <- RunPCA(WT, assay = "Spatial", reduction.name = "pca")
ElbowPlot(WT)

WT <- FindNeighbors(WT, reduction = "pca", dims = 1:14)

WT <- FindClusters(WT, cluster.name = "seurat_cluster", resolution = 1.7)

WT <- RunUMAP(WT, reduction = "pca", reduction.name = "umap", return.model = T, dims = 1:14)

DimPlot(WT, reduction = "umap")
```


#### Decomposition of celltypes
#### Merging public and own scRNAseq datasets 
```{r}
# load in the reference scRNA-seq dataset
ref <- mac_sc_wt
```


#first step: Change gene name format in skin_sc
```{r}
skin_sc <- readRDS(paste0(dir.s, "/Anderson.rds"))

table(skin_sc@meta.data$merged_cell_type_name)

table(skin_sc@meta.data$disease_ontology_term_id)

#KRT   FB CHFB  VEC  LEC  PVC  SCH   MC  LYM  NEU  APC 
#2642 3509  994  278  267  301  263   55  521  300  686

Idents(skin_sc) <- "disease_ontology_term_id"

skin_sc_he <- subset(skin_sc, idents = setdiff(levels(Idents(skin_sc)), "MONDO:0006541"))

Idents(skin_sc_he) <- "merged_cell_type_name"

skin_sc_he <- subset(skin_sc_he, idents = setdiff(levels(Idents(skin_sc_he)), "MC"))

levels(Idents(skin_sc_he))  

gene_symbols_pub <- rownames(GetAssayData(object = skin_sc_he, assay = "RNA", layer = "counts"))
head(gene_symbols_pub) 

gene_symbols_priv <- rownames(GetAssayData(object = ref, assay = "RNA", layer = "counts"))
head(gene_symbols_priv) 

# Step 1: Fetch the Mapping of Ensembl IDs to Gene Symbols
ensembl <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
original_genes <- rownames(GetAssayData(skin_sc_he, assay = "RNA", layer = "counts"))

mapping <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  filters = "ensembl_gene_id",
  values = original_genes,
  mart = ensembl
)

# Step 2: Map Ensembl IDs to Gene Symbols
gene_symbols <- mapping$external_gene_name[match(original_genes, mapping$ensembl_gene_id)]

# Handle unmapped genes by retaining original Ensembl IDs
gene_symbols[is.na(gene_symbols)] <- original_genes[is.na(gene_symbols)]

# Ensure all gene symbols are unique
gene_symbols <- make.unique(gene_symbols)

# Step 3: Replace Rownames in the Counts Matrix
counts_pub <- GetAssayData(object = skin_sc_he, assay = "RNA", layer = "counts")
rownames(counts_pub) <- gene_symbols

head(counts_pub)

sum(rownames(counts_pub) == "")  # Should return 0

rownames(counts_pub)[rownames(counts_pub) == ""] <- "Unnamed_Gene"

# Check for invalid rownames (e.g., empty or NA)
if (any(is.na(rownames(counts_pub))) || any(rownames(counts_pub) == "")) {
  stop("Counts matrix contains invalid feature names. Please check rownames.")
}

# Step 4: Replace the RNA Assay in the Seurat Object
skin_sc_he[["RNA"]] <- CreateAssayObject(counts = counts_pub)

# Step 5: Verify the Changes
head(rownames(GetAssayData(skin_sc_he, assay = "RNA", layer = "counts")))  # Check new gene names

dim(GetAssayData(skin_sc_he, assay = "RNA", layer = "counts"))  # Confirm dimensions remain unchanged

```


```{r}
# Normalize and find variable features for skin_sc_he
skin_sc_he <- NormalizeData(skin_sc_he)
skin_sc_he <- FindVariableFeatures(skin_sc_he, selection.method = "vst", nfeatures = 3000)

# Normalize and find variable features for ref
DefaultAssay(ref) <- "RNA"
ref <- NormalizeData(ref)
ref <- FindVariableFeatures(ref, selection.method = "vst", nfeatures = 3000)
```

#perform integration
```{r}
# Find integration anchors
anchors <- FindIntegrationAnchors(object.list = list(skin_sc_he, ref), dims = 1:30)
# Integrate the data
integrated <- IntegrateData(anchorset = anchors, dims = 1:30)

```

#annotate integrated
```{r}
integrated$ref_celltype <- NA  # Initialize with NA

# Transfer annotations from skin_sc_he to the integrated dataset
# This assumes that the cell names in skin_sc_he and integrated match (or can be subsetted/matched)
common_cells <- intersect(Cells(skin_sc_he), Cells(integrated))
integrated$ref_celltype[common_cells] <- skin_sc_he$merged_cell_type_name[common_cells]

# Assign "APC" to all cells originating from ref
# This assumes the cells from ref have unique identifiers or can be identified in integrated
ref_cells <- intersect(Cells(ref), Cells(integrated))
integrated$ref_celltype[ref_cells] <- "MAC"

table(integrated$ref_celltype)

# Create a named vector for mapping numeric labels to original names
label_mapping <- c(
  "1" = "KRT",
  "2" = "FB",
  "3" = "CHFB",
  "4" = "VEC",
  "5" = "LEC",
  "6" = "PVC",
  "7" = "SCH",
  "9" = "LYM",
  "10" = "NEU",
  "11" = "APC-MAC",
  "MAC" = "APC-MAC"
)

mapped_values <- label_mapping[integrated$ref_celltype]

table(mapped_values)

names(mapped_values) <- Cells(integrated)

# Add or update the ref_celltype metadata
integrated <- AddMetaData(integrated, metadata = mapped_values, col.name = "ref_celltype")

# Check the updated metadata
table(integrated@meta.data$ref_celltype)

Idents(integrated) <- "ref_celltype"

```

#visualize integrated
```{r}

names(integrated@meta.data)
table(integrated$disease)


set.seed(1)

# Scale the data and run PCA
DefaultAssay(integrated) <- "integrated"
integrated <- ScaleData(integrated, verbose = FALSE)
integrated <- RunPCA(integrated, npcs = 30, verbose = FALSE)

# Visualize PCA results
ElbowPlot(integrated, ndims = 30)


# Find neighbors and clusters
integrated <- FindNeighbors(integrated, dims = 1:30)
integrated <- FindClusters(integrated, resolution = 0.5)

# Run UMAP
integrated <- RunUMAP(integrated, dims = 1:30, seed.use = 1)

# Visualize UMAP
DimPlot(integrated, reduction = "umap", group.by = "ref_celltype")

# Check cluster composition
table(Idents(integrated))


table(integrated$Experiment, useNA = "ifany")
table(integrated$disease, useNA = "ifany")

integrated$Experiment[is.na(integrated$Experiment) & integrated$disease == "normal"] <- "whole_skin"

# Check if the update worked
table(integrated$Experiment, useNA = "ifany")


# If you want to compare between datasets
DimPlot(integrated, reduction = "umap", split.by = "Experiment", group.by = "ref_celltype")

saveRDS(integrated, paste0(dir.s, "/integrated.RDS"))

integrated <- readRDS(paste0(dir.s, "/integrated.RDS"))
```

#create ref dataset with integrated for RCTD:
```{r}
Idents(integrated) <- "ref_celltype"
levels(Idents(integrated))  

counts <- integrated[["RNA"]]$counts

cluster <- droplevels(Idents(integrated))

genes <- rownames(counts)

counts <- counts[!is.na(rownames(counts)), ]

anyDuplicated(rownames(counts))  # Returns the number of duplicates

nUMI <- colSums(counts)

# Create the RCTD reference object
reference <- Reference(counts, cluster, nUMI)

common_genes <- intersect(rownames(counts), rownames(reference@counts))
counts <- counts[common_genes, , drop = FALSE]
reference@counts <- reference@counts[common_genes, , drop = FALSE]
```

#perform RCTD with WT
```{r}
# Query preparation from the spatial dataset (WT)
counts_hd <- WT[["Spatial"]]$counts  # Spatial RNA counts
WT_cells_hd <- colnames(WT[["Spatial"]])  # Spatial cells
coords <- GetTissueCoordinates(WT)[WT_cells_hd, 1:2]  # Tissue coordinates for spatial data

# Create the RCTD query object
query <- SpatialRNA(coords, counts_hd, colSums(counts_hd))

# Run the RCTD algorithm
RCTD.wt <- create.RCTD(query, reference, max_cores = 7)
RCTD.wt@config$CELL_MIN <- 0
RCTD.wt <- run.RCTD(RCTD.wt, doublet_mode = "doublet")
```

# Save the RCTD results
```{r}
#saveRDS(RCTD.wt, paste0(dir.s, "/RCTD_WT_combined_ref.rds"))
RCTD.wt <- readRDS(paste0(dir.s, "/RCTD_WT_combined_ref.rds"))
```

#explore weights and apply filter
```{r}
# RCTD.wt <- readRDS(paste0(dir.s, "/RCTD_WT_combined_ref.rds"))

table(RCTD.wt@results$results_df$first_type)

head(RCTD.wt@results$weights)

results_df <- RCTD.wt@results$results_df
weights <- RCTD.wt@results$weights

# Extract weights for the primary cell type (first_type)
primary_weights <- sapply(1:nrow(results_df), function(i) {
    spot <- rownames(results_df)[i]
    cell_type <- results_df$first_type[i]
    weights[spot, cell_type]
})

# Add primary_weights as a separate DataFrame for exploration
primary_weights_df <- data.frame(
    spot = rownames(results_df),
    first_type = results_df$first_type,
    primary_weight = primary_weights
)

head(primary_weights_df)
nrow(primary_weights_df)

threshold <- 0.3  # Adjust as needed
filtered_primary_weights_df <- primary_weights_df[primary_weights_df$primary_weight >= threshold, ]

nrow(filtered_primary_weights_df)

# Filter results_df
filtered_results_df <- results_df[rownames(results_df) %in% filtered_primary_weights_df$spot, ]

# Filter weights
filtered_weights <- weights[rownames(weights) %in% filtered_primary_weights_df$spot, ]

table(filtered_results_df$first_type)

first_type_filtered <- filtered_results_df$first_type
names(first_type_filtered) <- rownames(filtered_results_df)

head(first_type_filtered)

# Add RCTD results back to the Seurat object
# Add original RCTD results
WT <- AddMetaData(WT, metadata = RCTD.wt@results$results_df)

# Add first_type_filtered as a new column
WT <- AddMetaData(WT, metadata = first_type_filtered, col.name = "first_type_filtered")

table(WT@meta.data$first_type_filtered)
table(WT@meta.data$first_type)

```

#subset data to include only spots for APC
```{r}
#subset for APC
WT$highlight_APC <- ifelse(WT$first_type == "APC-MAC", "APC-MAC", "Other")

WT$highlight_APC <- factor(WT$highlight_APC, levels = c("Other", "APC-MAC"))

subset_APC_wt <- subset(WT, subset = highlight_APC == "APC-MAC")

#subset for APC filtered
WT$highlight_APC_filtered <- ifelse(WT$first_type_filtered == "APC-MAC", "APC-MAC", "Other")

WT$highlight_APC_filtered <- factor(WT$highlight_APC_filtered, levels = c("Other", "APC-MAC"))

subset_APC_wt_filtered <- subset(WT, subset = highlight_APC_filtered == "APC-MAC")
```

# Perform differential expression analysis WT
```{r}
deg_results <- FindMarkers(WT, ident.1 = "APC-MAC", ident.2 = "Other", 
                           group.by = "highlight_APC", 
                           test.use = "wilcox", # Wilcoxon rank-sum test
                           logfc.threshold = 0.25, 
                           min.pct = 0.1) 

sorted_deg_results <- deg_results %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    arrange(desc(avg_log2FC))

# Export to CSV
write.csv(sorted_deg_results, file = paste0(dir.s,  "/APC_deg_wt.csv"), row.names = FALSE)


deg_results_filtered <- FindMarkers(WT, ident.1 = "APC-MAC", ident.2 = "Other", 
                           group.by = "highlight_APC_filtered", 
                           test.use = "wilcox", # Wilcoxon rank-sum test
                           logfc.threshold = 0.25, 
                           min.pct = 0.1) 

sorted_deg_results_filtered <- deg_results_filtered %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    arrange(desc(avg_log2FC))

# Export to CSV
write.csv(sorted_deg_results_filtered, file = paste0(dir.s,  "/APC_deg_wt_filtered.csv"), row.names = FALSE)

head(sorted_deg_results_filtered, n=20)
```

# Explore spot assignment WT
```{r}
table(WT@meta.data$first_type, useNA = "ifany")

SpatialPlot(WT, group.by = "first_type") +
    theme_minimal()


# SpatialPlot with manual color assignment
SpatialPlot(WT, group.by = "highlight_APC") +
    scale_fill_manual(values = c("Other" = "grey", "APC" = "red")) +
    theme_minimal()


VlnPlot(WT, features = "Csf1r", group.by = "highlight_APC", 
        cols = c("grey", "red")) +
    theme_minimal() +
    ggtitle("Expression of Csf1r in APC and Other Categories") +
    ylab("Expression Level") +
    xlab("Category")

gene_of_interest <- "Fcer1g"

# Extract the expression data for the gene
gene_expression <- FetchData(WT, vars = gene_of_interest)

# Add the corresponding categories to the data
gene_expression$highlight_APC <- WT$highlight_APC

# Calculate the percentage of zero values for each category
zero_percentage <- gene_expression %>%
    group_by(highlight_APC) %>%
    summarize(percentage_zero = mean(get(gene_of_interest) == 0) * 100)

# View the result
print(zero_percentage)
```



####Second step: Decomposition with own scRNA-seq data -> only APCs
#run decomposition with APC subset WT
```{r}
#check input data
table(ref@meta.data$broad_clusters)
Idents(ref) <- "broad_clusters"
levels(Idents(ref))

dim(subset_APC_wt_filtered)

#create ref parameters
counts <- ref[["RNA"]]$counts
cluster <- as.factor(ref$broad_clusters)
nUMI <- ref$nCount_RNA
cluster <- droplevels(cluster)

# create the RCTD reference objects
reference <- Reference(counts, cluster, nUMI)

#on APC_filtered
counts_hd <- subset_APC_wt_filtered[["Spatial"]]$counts
WT_cells_hd <- colnames(subset_APC_wt_filtered[["Spatial"]])
coords <- GetTissueCoordinates(subset_APC_wt_filtered)[WT_cells_hd, 1:2]

query.APC.wt <- SpatialRNA(coords, counts_hd, colSums(counts_hd))

RCTD.wt.APC <- create.RCTD(query.APC.wt, reference, max_cores = 7)
RCTD.wt.APC <- run.RCTD(RCTD.wt.APC, doublet_mode = "doublet")
```


```{r}
saveRDS(RCTD.wt.APC, paste0(dir.s, "/RCTD.wt.APC.rds"))

RCTD.wt.APC <- readRDS(paste0(dir.s, "/RCTD.wt.APC.rds"))
```

```{r}
#determine optimal filter threshold
## --- Inputs you already have -----------------------------------------------
results_df <- RCTD.wt.APC@results$results_df
weights    <- RCTD.wt.APC@results$weights

## Pull nUMI per spot (classic RCTD -> SpatialRNA; Bioc -> SpatialExperiment)
get_nUMI <- function(RCTD) {
  out <- NULL
  # classic spacexr object
  out <- tryCatch(RCTD@spatialRNA@nUMI, error = function(e) NULL)
  if (!is.null(out)) return(out)
  # Bioconductor version (stored in colData)
  out <- tryCatch(SummarizedExperiment::colData(RCTD@spatialRNA)$nUMI, error = function(e) NULL)
  if (!is.null(out)) { names(out) <- colnames(RCTD@spatialRNA); return(out) }
  stop("Could not locate nUMI; please supply a named numeric vector of nUMI per spot.")
}
nUMI_vec <- get_nUMI(RCTD.wt.APC)

## --- Compute primary weight -------------------------------------------------
primary_weights <- vapply(seq_len(nrow(results_df)), function(i) {
  spot <- rownames(results_df)[i]
  ct   <- results_df$first_type[i]
  weights[spot, ct]
}, numeric(1))

results_df$primary_weight <- primary_weights
results_df$nUMI           <- nUMI_vec[rownames(results_df)]

## --- Find a data-driven 'knee' in primary_weight ---------------------------
## Kneedle-style elbow on the sorted primary_weight ECDF
find_knee <- function(w) {
  w <- sort(w)
  n <- length(w)
  if (n < 10) return(NA_real_)
  x <- seq_len(n) / n
  # line through (x1,w1) and (xn,wn) ~ (1/n, min) to (1, max)
  x1 <- x[1]; y1 <- w[1]; xN <- x[n]; yN <- w[n]
  # distance from each point to the line
  # line ax + by + c = 0
  a <- (y1 - yN)
  b <- (xN - x1)
  c <- (x1*yN - xN*y1)
  # normalized perpendicular distance
  d <- abs(a*x + b*w + c) / sqrt(a^2 + b^2)
  knee_idx <- which.max(d)
  knee_val <- w[knee_idx]
  knee_val
}
knee <- find_knee(results_df$primary_weight)

results_df_plot <- subset(results_df, !is.na(primary_weight))

# 1) Histogram with knee and shaded kept area
p1 <- ggplot(results_df_plot, aes(x = primary_weight)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = knee, linetype = "dashed", linewidth = 0.8) +
  annotate("rect", xmin = knee, xmax = Inf, ymin = -Inf, ymax = Inf, alpha = 0.08) +
  labs(
    title = "Primary weight distribution",
    x = "primary_weight", y = "count",
    subtitle = sprintf("Global knee = %.3f (kept region shaded)", knee)
  )

# 2) Sorted curve with knee point and endpoint line (the “knee” geometry)
w_sorted <- sort(results_df_plot$primary_weight)
n <- length(w_sorted)
df_sorted <- data.frame(x = seq_len(n) / n, y = w_sorted)

# Line through endpoints of the sorted curve
line_y <- function(x) {
  y1 <- w_sorted[1]; yN <- w_sorted[n]
  x1 <- 1 / n;       xN <- 1
  y1 + (yN - y1) * (x - x1) / (xN - x1)
}
df_sorted$y_line <- line_y(df_sorted$x)

# Knee row (closest y to knee on the sorted curve) for highlighting
knee_idx <- which.min(abs(df_sorted$y - knee))
knee_pt  <- df_sorted[knee_idx, ]

p2 <- ggplot(df_sorted, aes(x, y)) +
  geom_line() +
  geom_line(aes(y = y_line), linetype = "dotted") +
  geom_point(data = knee_pt, size = 2) +
  geom_segment(aes(x = knee_pt$x, xend = knee_pt$x, y = knee_pt$y, yend = knee),
               linetype = "dashed") +
  geom_hline(yintercept = knee, linetype = "dashed") +
  labs(
    title = "Sorted primary_weight with knee",
    subtitle = "Solid: weights · Dotted: endpoint line · Point: knee",
    x = "ECDF x (rank / n)", y = "primary_weight"
  )

# Print both
print(p1)
print(p2)

# Quick summary
keep_mask <- results_df$primary_weight >= knee
cat(sprintf("\nKept %d / %d spots (%.1f%%) with primary_weight ≥ %.3f\n",
            sum(keep_mask, na.rm = TRUE),
            nrow(results_df),
            100 * mean(keep_mask, na.rm = TRUE),
            knee))

# Histogram
ggplot2::ggsave(filename = file.path(dir.s, "primary_weight_hist.png"),
                plot = p1, width = 7, height = 5, dpi = 300)

# Sorted curve
ggplot2::ggsave(filename = file.path(dir.s, "primary_weight_sorted_knee.png"),
                plot = p2, width = 7, height = 5, dpi = 300)
```



```{r}
# Extract weights and results_df
results_df <- RCTD.wt.APC@results$results_df
weights <- RCTD.wt.APC@results$weights

# Extract primary weights for first_type
primary_weights <- sapply(1:nrow(results_df), function(i) {
    spot <- rownames(results_df)[i]
    cell_type <- results_df$first_type[i]
    weights[spot, cell_type]
})

# Create primary_weights_df
primary_weights_df <- data.frame(
    spot = rownames(results_df),
    first_type = results_df$first_type,
    primary_weight = primary_weights
)

# Filter spots based on the threshold
threshold <- knee  # Adjust as needed
filtered_primary_weights_df <- primary_weights_df[primary_weights_df$primary_weight >= threshold, ]

# Filter results_df for spots passing the threshold
filtered_results_df <- results_df[rownames(results_df) %in% filtered_primary_weights_df$spot, ]

# Create named vector for mac_type_filtered
mac_type_filtered <- filtered_results_df$first_type
names(mac_type_filtered) <- rownames(filtered_results_df)

#create full metadata table
full_MAC <- rep(NA, ncol(WT))
names(full_MAC) <- colnames(WT)

# Fill in the values for cells that are present in mac_type_filtered
full_MAC[names(mac_type_filtered)] <- mac_type_filtered

table(full_MAC)

# Now add the complete metadata vector
WT <- AddMetaData(object = WT, metadata = full_MAC, col.name = "MAC_type_filtered")


# Add original first_type as MAC_type
MAC_type <- results_df$first_type
names(MAC_type) <- rownames(results_df)

head(full_MAC)

#create full metadata table
full_MAC <- rep(NA, ncol(WT))
names(full_MAC) <- colnames(WT)

full_MAC[names(MAC_type)] <- MAC_type

WT <- AddMetaData(WT, metadata = full_MAC, col.name = "MAC_type")

# Confirm metadata addition
table(WT@meta.data$MAC_type_filtered, useNA = "ifany")
table(WT@meta.data$MAC_type, useNA = "ifany")
```

#transfer spatial data from previous object
```{r}
WT_previous <- readRDS(paste0(dir.s, "/WT.spatial8.rds"))
names(WT_previous@meta.data)


# Extract the desired metadata columns from WT_previous
metadata_to_transfer <- WT_previous@meta.data[, c("distance_to_border_um", "distance_to_hb_um")]

dim(metadata_to_transfer)

# Add the metadata to the WT object
WT <- AddMetaData(WT, metadata = metadata_to_transfer)

names(WT@meta.data)

metadata_to_export <- WT@meta.data[, c("distance_to_border_um", "distance_to_hb_um", "MAC_type_filtered", "MAC_type", "first_type", "first_type_filtered")]

dim(metadata_to_export)
colnames(metadata_to_export)
table(metadata_to_export$first_type_filtered, useNA = "ifany")


write.csv(metadata_to_export, file = paste0(dir.s, "/WT_metadata_export.csv"), row.names = TRUE)


```



```{r}
saveRDS(WT, paste0(dir.s, "/WT.spatial_annotated.rds"))
```




