---
title: "Gene set enrichment"
output: html_document
date: '2022-09-28'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(clusterProfiler)
library(org.Mm.eg.db)
library(GO.db)
library(Seurat)
library(tidyverse)
library(cowplot)
library(patchwork)
library(presto)
library(pheatmap)
library(RColorBrewer)

dir.m <- "Results/gene set enrichment/markers"
dir.p <- "Results/gene set enrichment/plots"
dir.g <- "Results/gene set enrichment/GSEA"

# Define colors for clusters 0 to 11
colors <- c(
  "0" = "#a6cee3", "1" = "#1f78b4", "2" = "#b2df8a", "3" = "#33a02c",
  "4" = "#e31a1c", "5" = "#ff7f00", "6" = "#cab2d6", "7" = "#6a3d9a",
  "8" = "#ffff99", "9" = "#b15928", "10" = "#fb9a99", "11" = "#005F6A"
)
```


```{r setup, include=FALSE}
mac <- readRDS("Results/2_Integration/mac.RDS")
DefaultAssay(mac) <- "RNA"
mac<- NormalizeData(mac, normalization.method = "LogNormalize") 
steady <- subset(x = mac, subset = treatment=="steady")
infected <- subset(x = mac, subset = treatment=="infected")
wt <- subset(x = mac, subset = genotype=="WT")
irf8 <- subset(x = mac, subset = genotype=="Irf8ko")
inf.wt<- subset(x = infected, subset = genotype=="WT")
inf.irf<- subset(x = infected, subset = genotype=="Irf8ko")

WT_steady.markers<- read_excel("Results/3FindMarkers/A.markers.xlsx")
WT_infected.markers<- read_excel("Results/3FindMarkers/B.markers.xlsx")
Irf_steady.markers<- read_excel("Results/3FindMarkers/C.markers.xlsx")
Irf_infected.markers<- read_excel("Results/3FindMarkers/D.markers.xlsx")



```

#A: WT-steady
#B: WT-infected
#C: Irf-steady
#D: Irf-infected

#find_and_save_markers
```{r}

find_and_save_markers <- function(obj_name, num_clusters = 11) {
  # Loop through each cluster from 0 to num_clusters
  for (i in 0:num_clusters) {
    # Generate the cluster-specific object name for saving
    cluster_name <- paste0(obj_name,"_cl", i)
    
    # Find markers for the current cluster
    markers <- FindMarkers(get(obj_name), ident.1 = i, logfc.threshold = 0.25)
    
    # Save the marker data as an RDS file in the specified directory
    saveRDS(markers, file = paste0(dir.m, "/", cluster_name, ".RDS"))
  }
}

find_and_save_markers("inf.wt")
find_and_save_markers("inf.irf")
```

#perform_gsea for each cluster
```{r}
perform_gsea <- function(obj_name, orgdb, num_clusters = 11) {
  
  # Inner function to perform GSEA for a single cluster
  perform_gsea_for_cluster <- function(cluster_data, cluster_id) {
    set.seed(123)  # For reproducibility
    
    # Move row names to a column called 'gene'
    cluster_data$gene <- rownames(cluster_data)
    rownames(cluster_data) <- NULL
    
    # Convert gene symbols to ENTREZ IDs
    eg <- bitr(cluster_data$gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = orgdb)
    eg <- subset(eg, !duplicated(eg$SYMBOL))
    rownames(eg) <- eg$SYMBOL
    eg$SYMBOL <- NULL
    
    # Merge with marker data to get relevant ENTREZ IDs
    diffexpgenes <- merge(cluster_data, eg, by.x = "gene", by.y = "row.names")
    rownames(diffexpgenes) <- diffexpgenes$gene
    diffexpgenes$gene <- NULL
    
    # Prepare gene list for GSEA
    fc <- diffexpgenes$avg_log2FC
    names(fc) <- diffexpgenes$ENTREZID
    fc <- sort(fc, decreasing = TRUE)
    
    # Perform GSEA analysis for Biological Process (BP) GO terms only
    gse <- gseGO(geneList = fc, ont = "BP", minGSSize = 15, maxGSSize = 800, OrgDb = orgdb, seed = 1)
    
    # Check if GSEA result is available
    if (is.null(gse) || nrow(gse@result) == 0) {
      message(paste("No significant GO terms for cluster", cluster_id))
      return(NULL)
    }
    
    # Sort results by enrichmentScore
    sorted_results <- gse@result[order(gse@result$enrichmentScore, decreasing = TRUE), ]
    
    # Save sorted results to CSV and RDS
    csv_output <- file.path(dir.g, paste0("gse_", obj_name, "_cl", cluster_id, ".csv"))
    rds_output <- file.path(dir.g, paste0("gse_", obj_name, "_cl", cluster_id, ".rds"))
    
    write.csv(sorted_results, csv_output, row.names = FALSE)
    saveRDS(gse, rds_output)
    
    message(paste("Sorted GSEA results saved for cluster", cluster_id, "at", csv_output, "and", rds_output))
  }
  
  # Loop through each cluster RDS file
  for (i in 0:num_clusters) {
    # Construct the file path for the marker data
    rds_file <- file.path(dir.m, paste0(obj_name, "_cl", i, ".RDS"))
    
    # Load the marker data for the cluster
    if (!file.exists(rds_file)) {
      message(paste("Marker file not found:", rds_file))
      next
    }
    cluster_data <- readRDS(rds_file)
    
    # Ensure cluster_data has necessary data
    if (nrow(cluster_data) == 0) {
      message(paste("No data available in cluster", i))
      next
    }
    
    # Perform GSEA for the current cluster
    perform_gsea_for_cluster(cluster_data, i)
  }
}


# usage:
perform_gsea("inf.wt", org.Mm.eg.db)
perform_gsea("inf.irf", org.Mm.eg.db)
```

#combine all into one file WT
```{r}
# Define clusters
clusters <- c(0,1,2,6,7,8)
results <- list()

# Read and process results
for (cl in clusters) {
  gsea_file <- paste0(dir.g, "/gse_inf.wt_cl", cl, ".csv")
  res <- read.csv(gsea_file)
  res$Cluster <- paste0("Cluster_", cl)
  results[[cl + 1]] <- res
}

# Combine all results
combined_wt <- do.call(rbind, results)
head(combined_wt)
```

#combine all into one file Irf8ko
```{r}
# Define clusters
clusters <- c(1,3,4,5,6,7,8,9)
results <- list()

# Read and process results
for (cl in clusters) {
  gsea_file <- paste0(dir.g, "/gse_inf.irf_cl", cl, ".csv")
  res <- read.csv(gsea_file)
  res$Cluster <- paste0("Cluster_", cl)
  results[[cl + 1]] <- res
}

# Combine all results
combined_irf <- do.call(rbind, results)
```

# Filter for manually selected GO terms
```{r}
selected_go_terms <- c("GO:0015732", "GO:0006801", "GO:0071456", "GO:0002526",   #0
                       "GO:0070741", "GO:0007015", #1
                       "GO:1903901", "GO:0019882", "GO:0071357",  "GO:0140374", "GO:0034341", "GO:0045087", #2
                      
                       "GO:0032456", #6
                       "GO:0009581", "GO:0006695", "GO:0002694", #7
                       "GO:0009063"#8
                       ) 
```

#extra terms for Irf8?
```{r}
irf_go_terms <- c(   "GO:0070741", "GO:0035458", #1
                       "GO:0017004", "GO:0033108",  #3
                       "GO:0045292", #4
                       "GO:0046365", "GO:0019884", "GO:0001912",  #5
                     #6
                    "GO:0002717", "GO:0042267", "GO:0001910", "GO:1903039",  #7
                    "GO:0007031", "GO:0032365",   #8
                     "GO:0018149" #9
                       ) 
```


#create heatmap WT
```{r}
filtered_wt <- combined_wt[combined_wt$ID %in% selected_go_terms, ]
head(filtered_wt)

heatmap_data <- reshape2::dcast(
  filtered_wt,
  ID + Description ~ Cluster,
  value.var = "NES"
)

head(heatmap_data)


heatmap_data <- heatmap_data[match(selected_go_terms, heatmap_data$ID), ]
###before capitalization we need to reorder heatmap_data

capitalize_first_letter <- function(term) {
  sapply(term, function(x) paste0(toupper(substring(x, 1, 1)), substring(x, 2)))
}

heatmap_data$Description <- capitalize_first_letter(heatmap_data$Description)
head(heatmap_data)

rownames(heatmap_data) <- heatmap_data$Description

colnames(heatmap_data) <- gsub("Cluster_", "Cluster ", colnames(heatmap_data))

heatmap_data <- heatmap_data[, -c(1:2)]  # Remove the Description column

color_palette <- rev(colorRampPalette(brewer.pal(11, "RdYlBu"))(100))

png(filename = paste0(dir.p, "/GO_Term_Heatmap_wt.png"), width = 6, height = 3.7, units = "in", res = 600)

# Generate heatmap
pheatmap(
  heatmap_data,
  cluster_rows = FALSE,       
  cluster_cols = FALSE,       
  color = color_palette,      
  main = "",
  fontsize = 10,
  labels_col = colnames(heatmap_data),
)

dev.off()
```


#create heatmap Irf8
```{r}
filtered_irf <- combined_irf[combined_irf$ID %in% irf_go_terms, ]
head(filtered_irf)

heatmap_data <- reshape2::dcast(
  filtered_irf,
  ID + Description ~ Cluster,
  value.var = "NES"
)

head(heatmap_data)


heatmap_data <- heatmap_data[match(irf_go_terms, heatmap_data$ID), ]
###before capitalization we need to reorder heatmap_data

capitalize_first_letter <- function(term) {
  sapply(term, function(x) paste0(toupper(substring(x, 1, 1)), substring(x, 2)))
}

heatmap_data$Description <- capitalize_first_letter(heatmap_data$Description)
head(heatmap_data)

rownames(heatmap_data) <- heatmap_data$Description

colnames(heatmap_data) <- gsub("Cluster_", "Cluster ", colnames(heatmap_data))

heatmap_data <- heatmap_data[, -c(1:2)]  # Remove the Description column

color_palette <- rev(colorRampPalette(brewer.pal(11, "RdYlBu"))(100))

png(filename = paste0(dir.p, "/GO_Term_Heatmap_irf.png"), width = 6, height = 3.7, units = "in", res = 600)

# Generate heatmap
pheatmap(
  heatmap_data,
  cluster_rows = FALSE,       
  cluster_cols = FALSE,       
  color = color_palette,      
  main = "",
  fontsize = 10,
  labels_col = colnames(heatmap_data),
)

dev.off()
```




####Exploration starts here



#load res-mac files
```{r}
folder_path <- "Results/3FindMarkers/Res_infected_comp"

WT_res.cl7<- read_excel("Results/3FindMarkers/Res_infected_comp/WT_res.cl7.xlsx")
WT_res.cl1<- read_excel("Results/3FindMarkers/Res_infected_comp/WT_res.cl1.xlsx")
WT_res.cl8<- read_excel("Results/3FindMarkers/Res_infected_comp/WT_res.cl8.xlsx")
WT_res.cl10<- read_excel("Results/3FindMarkers/Res_infected_comp/WT_res.cl10.xlsx")

Irf_res.cl7<- read_excel("Results/3FindMarkers/Res_infected_comp/Irf_res.cl7.xlsx")
Irf_res.cl1<- read_excel("Results/3FindMarkers/Res_infected_comp/Irf_res.cl1.xlsx")
Irf_res.cl8<- read_excel("Results/3FindMarkers/Res_infected_comp/Irf_res.cl8.xlsx")
Irf_res.cl10<- read_excel("Results/3FindMarkers/Res_infected_comp/Irf_res.cl10.xlsx")
```



```{r}
plot_gsea_results <- function(condition, cluster_number, show_categories, colors, title = NULL) {
  
  # Construct the file path for the cluster-specific GSEA results
  gse_file <- file.path(dir.g, paste0("gse_", condition, "_cl", cluster_number, ".rds"))
  
  # Load the GSEA results
  gse_results <- readRDS(gse_file)
  
  # Get the color for the current cluster
  cluster_color <- colors[[as.character(cluster_number)]]
  
  plot_title <-  title 
  
  # capitalize the first letter and break labels after a max number of words
  format_label <- function(label, max_words) {
    label <- paste0(toupper(substr(label, 1, 1)), substr(label, 2, nchar(label)))
    words <- unlist(strsplit(label, " "))
    
    # Add line breaks if label exceeds max words
    if (length(words) > max_words) {
      lines <- split(words, ceiling(seq_along(words) / max_words))  
      label <- paste(sapply(lines, paste, collapse = " "), collapse = "\n")  
    }
    
    return(label)
  }
  
  # Plot the GSEA results using dotplot
  plot <- dotplot(gse_results, showCategory = show_categories, split = ".sign", font = 14, title = plot_title) +
    facet_grid(. ~ .sign) + 
    scale_colour_gradientn(colours = c("red4", "cornflowerblue")) +
    theme(
      plot.title = element_text(size = 16, colour = cluster_color),
      axis.text.y = element_text(size = 14),  
      axis.text.x = element_text(angle = 45, hjust = 1, size = 10)  
    )
  
  # Modify the category labels
  plot <- plot + scale_y_discrete(labels = function(labels) {
    sapply(labels, function(label) format_label(label, max_words = 4))
  })
  
  # Save the plot as a PNG file
  output_file <- file.path(dir.p, paste0("GSEAdot_", condition, "_Cl", cluster_number, ".png"))
  ggsave(output_file, plot = plot, height = 3.5, width = 6.5, dpi = 600)
}

```



#load display and select terms
```{r}
gse_inf_wt_cl0 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.wt_cl0.csv")
gse_inf_wt_cl0 <- gse_inf_wt_cl0[order(gse_inf_wt_cl0$enrichmentScore, decreasing = TRUE), ]
gse_inf_wt_cl0$Description

gse_inf_wt_cl2 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.wt_cl2.csv")
gse_inf_wt_cl2 <- gse_inf_wt_cl2[order(gse_inf_wt_cl2$enrichmentScore, decreasing = TRUE), ]
gse_inf_wt_cl2$Description

gse_inf_wt_cl7 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.wt_cl7.csv")
gse_inf_wt_cl7 <- gse_inf_wt_cl7[order(gse_inf_wt_cl7$enrichmentScore, decreasing = TRUE), ]
gse_inf_wt_cl7$Description

```

# Create plots-WT:
```{r}
plot_gsea_results(
   condition = "inf.wt",
   cluster_number = 0,
   show_categories = c("acute inflammatory response", "cellular response to hypoxia", "nitric oxide metabolic process", "extracellular matrix assembly", "collagen fibril organization", "regulation of fatty acid transport", "regulation of transforming growth factor beta production"),
   colors = colors,
   title = "WT Cluster 0 - infected"
 )

plot_gsea_results(
   condition = "inf.wt",
   cluster_number = 2,
   show_categories = c("antigen processing and presentation", "response to type I interferon", "defense response to virus", "response to type II interferon", "adaptive immune response", "extracellular matrix assembly", "tissue regeneration"),
   colors = colors,
   title = "WT Cluster 2 - infected"
 )

plot_gsea_results(
   condition = "inf.wt",
   cluster_number = 7,
   show_categories = c("arachidonic acid metabolic process",  "cholesterol biosynthetic process", "innate immune response",  "cytokine production", "response to oxidative stress",  "response to type I interferon"),
   colors = colors,
   title = "WT Cluster 7 - infected"
 )

```

#load display and select terms

```{r}
gse_inf_irf_cl1 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.irf_cl1.csv")
gse_inf_irf_cl1 <- gse_inf_irf_cl1[order(gse_inf_irf_cl1$enrichmentScore, decreasing = TRUE), ]
gse_inf_irf_cl1$Description

gse_inf_irf_cl3 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.irf_cl3.csv")
gse_inf_irf_cl3 <- gse_inf_irf_cl3[order(gse_inf_irf_cl3$enrichmentScore, decreasing = TRUE), ]
gse_inf_irf_cl3$Description

gse_inf_irf_cl4 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.irf_cl4.csv")
gse_inf_irf_cl4 <- gse_inf_irf_cl4[order(gse_inf_irf_cl4$enrichmentScore, decreasing = TRUE), ]
gse_inf_irf_cl4$Description

gse_inf_irf_cl5 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.irf_cl5.csv")
gse_inf_irf_cl5 <- gse_inf_irf_cl5[order(gse_inf_irf_cl5$enrichmentScore, decreasing = TRUE), ]
gse_inf_irf_cl5$Description

gse_inf_irf_cl10 <- read_csv("Results/gene set enrichment/GSEA/gse_inf.irf_cl10.csv")
gse_inf_irf_cl10 <- gse_inf_irf_cl10[order(gse_inf_irf_cl10$enrichmentScore, decreasing = TRUE), ]
gse_inf_irf_cl10$Description

#5,7,9,10
```


# Create plots-Irf8:
```{r}
plot_gsea_results(
   condition = "inf.irf",
   cluster_number = 1,
   show_categories = c("cellular response to interleukin-6","cellular response to lipopolysaccharide", "cell chemotaxis", "skin development", "acute inflammatory response", "integrin-mediated signaling pathway"),
   colors = colors,
   title = bquote(italic("Irf8")^"-/-" ~ "Cluster 1 - infected")
 )

plot_gsea_results(
   condition = "inf.irf",
   cluster_number = 3,
   show_categories = c("monosaccharide catabolic process", "oxidative phosphorylation","myeloid leukocyte activation", "positive regulation of defense response", "activation of innate immune response", "response to hypoxia", "regulation of cytokine production"),
   colors = colors,
   title = bquote(italic("Irf8")^"-/-" ~ "Cluster 3 - infected")
 )

plot_gsea_results(
   condition = "inf.irf",
   cluster_number = 4,
   show_categories = c("positive regulation of cell adhesion",  "regulation of anatomical structure morphogenesis", "response to wounding", "granulocyte activation", "actin cytoskeleton organization"),
   colors = colors,
   title = bquote(italic("Irf8")^"-/-" ~ "Cluster 4 - infected")
 )

plot_gsea_results(
   condition = "inf.irf",
   cluster_number = 5,
   show_categories = c("antigen processing and presentation of exogenous peptide antigen","regulation of natural killer cell mediated immunity", "oxidative phosphorylation", "regulation of immune response", "innate immune response", "toll-like receptor signaling pathway"),
   colors = colors,
   title = bquote(italic("Irf8")^"-/-" ~ "Cluster 5 - infected")
 )

plot_gsea_results(
   condition = "inf.irf",
   cluster_number = 7,
   show_categories = c("antigen processing and presentation of exogenous antigen", "natural killer cell mediated immunity", "immune response", "tissue development", "inflammatory response","extracellular matrix assembly"),
   colors = colors,
   title = bquote(italic("Irf8")^"-/-" ~ "Cluster 7 - infected")
 )


```

